Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-14 paper/paper.tex
grammar fixes

  Kiselyov's methods for implicit and explicit sharing detection for finally
  tagless style DSLs is an elegant solution without having to resort to unsafe
  observable sharing. However these methods are not applicable to all types of
  DSLs (including those generating hypergraphs). We will present alternative
  methods which handle these cases. The main difference comes from the use of a
  trie to perform hashconsing. Our method for implicit sharing essentially
  trades worst-case exponential growth in computation for increased memory
  footprint. To mitigate this issue, our method for explicit sharing reduces the
  memory footprint.

@@

\vspace{-1cm}
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/hashconscmp.png}
  \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the
    number of \mintinline{haskell}{add} operations performed.
    Hashcons is performed without explicit sharing and is clearly exponential,
    Triecons (without explicit sharing) and HashCons Explicit (with explicit
    sharing) overlap and are both linear
  } \label{fig:hashcons}
\end{figure}
As shown in Fig.~\ref{fig:hashcons}, this code will perform approximately
$2^{n+1}$ \mintinline{haskell}{hashcons} operations, where $n$ is the number of \mintinline{haskell}{add}s.

@@

The issue is that DAG generation requires splitting the state monad in two:
\begin{minted}{haskell}
instance Exp Graph where
  ...
  novel e1 e2 = let
     g1 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel1 h1 h2)
     g2 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel2 h1 h2)
     in (g1,g2)
\end{minted}
Each output it returns will now have to be individually evaluated, so a chain of
DSL functions that output 2 or more values will suffer from the same exponential
explosion of hashcons operations, and trying to adapt the let construct above,
just creates another function with the same problem (multiple outputs).

@@

This will prevent
redundant hashconsing without the need for explicit sharing,
but at the expense of storing redundant byte strings.

@@

\begin{table}
  \begin{tabular}{|l|l|l|l|l|l|l|l|}
    \hline
    {\bf Size } &  150 & 200 & 10000 & 50000 \\
    \hline
    {\bf HashCons time } &  0.0 secs & 0.0 secs & 0.01 secs & 0.03 secs \\
    {\bf HashCons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
    {\bf TrieCons time } &   0.0 secs & 0.0 secs & 0.03 secs & 0.16 secs \\
    {\bf TrieCons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
    \hline
  \end{tabular}

@@

We used this language to implement approximations of vector cos and tan (i.e., a
loop body that computes over an input and output array). The loop body was
unrolled by a factor of four making for a sufficiently large resulting DAG.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
  \caption{Comparison of cos using hashcons vs triecons}
  \label{fig:hashvscons}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of tan using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}

>>>>

  Kiselyov's methods for implicit and explicit sharing detection for finally
  tagless style DSLs is an elegant solution without having to resort to unsafe
  observable sharing. However these methods are not applicable to all types of
  DSLs (including those generating hypergraphs). We will present alternative
  methods which handle these cases. The main difference comes from the use of a
  trie to perform hash-consing. Our method for implicit sharing essentially
  trades worst-case exponential growth in computation for increased memory
  footprint. To mitigate this issue, our method for explicit sharing reduces the
  memory footprint.

@@

\vspace{-1cm}
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/hashconscmp.png}
  \caption{Number of calls to \mintinline{haskell}{hashcons} plotted against the
    number of \mintinline{haskell}{add} operations performed.
    Hash-consing is performed without explicit sharing and is clearly exponential,
    Triecons (without explicit sharing) and Hashcons Explicit (with explicit
    sharing) overlap and are both linear
  } \label{fig:hashcons}
\end{figure}
As shown in Fig.~\ref{fig:hashcons}, this code will perform approximately
$2^{n+1}$ \mintinline{haskell}{hashcons} operations, where $n$ is the number of \mintinline{haskell}{add}s.

@@

The issue is that DAG generation requires splitting the state monad in two:
\begin{minted}{haskell}
instance Exp Graph where
  ...
  novel e1 e2 = let
     g1 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel1 h1 h2)
     g2 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel2 h1 h2)
     in (g1,g2)
\end{minted}
Each output it returns will now have to be individually evaluated, so a chain of
DSL functions that output 2 or more values will suffer from the same exponential
explosion of hash-consing operations, and trying to adapt the let construct above,
just creates another function with the same problem (multiple outputs).

@@

This will prevent
redundant hash-consing without the need for explicit sharing,
but at the expense of storing redundant byte strings.

@@

\begin{table}
  \begin{tabular}{|l|l|l|l|l|l|l|l|}
    \hline
    {\bf Size } &  150 & 200 & 10000 & 50000 \\
    \hline
    {\bf Hash-Cons time } &  0.0 secs & 0.0 secs & 0.01 secs & 0.03 secs \\
    {\bf Hash-Cons alloc } &  619,296 bytes & 739,304 bytes & 28,662,592 bytes & 155,993,544 bytes \\
    {\bf Trie-Cons time } &   0.0 secs & 0.0 secs & 0.03 secs & 0.16 secs \\
    {\bf Trie-Cons alloc } &  1,773,416 bytes & 2,333,680 bytes & 129,146,808 bytes & 723,437,504 bytes \\
    \hline
  \end{tabular}

@@

We used this language to implement approximations of vector \texttt{cos} and \texttt{tan} (i.e., a
loop body that computes over an input and output array). The loop body was
unrolled by a factor of four making for a sufficiently large resulting DAG.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
  \caption{Comparison of cos using hash-cons vs trie-cons}
  \label{fig:hashvscons}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}
